---
title: "Getting Started with PetfindeR"
author: "Aaron Schlegel"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with PetfindeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

The goal of the PetfindeR package is to provide a simple and straightforward interface for interacting with the [Petfinder API](https://www.petfinder.com/developers/api-docs) through R. The Petfinder database contains approximately 300,000 adoptable pet records and 11,000 animal welfare organization records, which makes it a very useful and valuable source of data for those in the animal welfare community. However, the outputs from the Petfinder API are in messy JSON format and thus it makes it more time-consuming and often frustrating to coerce the output data into a format that is workable with R. The PetfindeR package was developed to alleviate this difficultly and provide users with a streamlined interface for getting the data they want and using it for their specific purpose.

# Obtaining an API Key from Petfinder

Before we can use the PetfindeR package, we need to obtain an API key from the Petfinder website to authenticate our requests. To acquire an API key, create an account on [Petfinder's developer page](https://www.petfinder.com/developers/api-key). A 'secret' key is also given from Petfinder for requests that require an additional layer of authentication; however, the current methods available do not require a secret key to be given (this could potentially change in the future).

# Getting Started with PetfindeR

```{r}
library(PetfindeR)
```

After installing and loading the package, The first step in using the PetfindeR package is to initialize the API connection to Petfinder. To do this, we pass the API key received from Petfinder into the `Petfinder` function which calls a wrapped [R6 class](https://cran.r-project.org/web/packages/R6/) and creates the connection to the API.

It is best practice to obfuscate sensitive data such as API keys to avoid any potential malicious activity. To this effect, I load my API key by using the `Sys.getenv()` function to access an environment variable containing the key.  

```{r}
key <- Sys.getenv('PETFINDER_KEY')
```

The API key is securely loaded, so we can initalize the PetfindeR API!

```{r}
pf <- Petfinder(key)
```

That's all there is to it! The [Petfinder API methods](https://www.petfinder.com/developers/api-docs#methods) are now accessible through R.

# Examples 

The following section introduces the `PetfindeR` package and how to interact with the Petfinder API with some simple examples on extracting data from the database. The examples are split into pet and shelter methods, not due to any significant demarcation, but rather for organization.  

## Pet Methods

| Method                | Petfinder API Method | Description                                                                                        |
|-----------------------|----------------------|----------------------------------------------------------------------------------------------------|
| breed.list()          | breed.list           | Returns the available breeds for the selected animal.                                              |
| pet.find()            | pet.find             | Returns a collection of pet records matching input parameters.                                     |
| pet.get()             | pet.get              | Returns a single record for a pet.                                                                 |
| pet.getRandom()       | pet.getRandom        | Returns a randomly selected pet record. The possible result can be filtered with input parameters. |

### Finding the available breeds of an animal

The `breed.list()` method allows the user to pull a list of breeds associated with a particular type of animal. The currently available animals to search for are `'barnyard'`, `'bird'`, `'cat'`, `'dog'`, `'horse'`, `'reptile'` and `'smallfurry'`. Let's say we are interested in finding the breeds of cats in the Petfinder database.

```{r}
cats <- pf$breed.list('cat')
```

```{r}
```

### Finding pet records matching criteria

The `pet.find()` method enables one to return a `data.frame` of pet records that match the input parameters. The available parameters can be found using `?pet.find()` or checking out the [Petfinder API documentation](https://www.petfinder.com/developers/api-docs#methods) directly.

Let's say we want to find female cats in Washington. The default amount of records is 25 (assuming there are 25 to return), which can be overridden with the `count` parameter.

```{r}
wa_cats <- pf$pet.find('WA', 'cat', sex='F')
```

### Returning random pet records

The `pet.getRandom()` method returns a random pet ID by default, but can return a full pet record with a description by setting `output = 'full'`. 

```{r}
pf$pet.getRandom(output = 'full')
```

We can also get a set of random results by setting the `records` parameter to the number of desired results to return. Note that each record counts as one call to the Petfinder API.

```{r}
random_records <- pf$pet.getRandom(records = 3, output = 'full')
random_records
```

### Getting pet records associated with a petId

Given a petId, an ID associated with information on a pet in the Petfinder database, the `pet.get()` method will return the pet data in a `data.frame`. The method also accepts a vector or list of IDs for returning multiple records. For example, we can take the petIds given in our previous call to `pet.getRandom()` and pass those to `pet.get()`. Note since we are using the same petIds from before, the returned records will be exactly the same.

## Shelter Methods

| Method                | Petfinder API Method | Description                                                                                        |
|-----------------------|----------------------|----------------------------------------------------------------------------------------------------|
| shelter.find()        | shelter.find         | Returns a collection of shelter records matching input parameters.                                 |
| shelter.get()         | shelter.get          | Returns a single shelter record.                                                                   |
| shelter.getPets()     | shelter.getPets      | Returns a collection of pet records for an individual shelter.                                     |
| shelter.listByBreed() | shelter.listByBreed  | Returns a list of shelter IDs listing animals matching the input animal breed.                     |

Shelter methods are rather similar to the Pet methods we explored previously, but return information on animal shelters and other animal welfare organizations.

### Finding animal welfare organizations in an area

Using the `shelter.find()` method, one can locate shelters and their available attributes matching the input parameters. For example, say we wanted to find shelters in a ZIP code in Seattle (my old zip code).

```{r}
wa_shelters <- pf$shelter.find('98115')
```

### Returning information on specific AWOs

Similar to `pet.get()`, the `shelter.get()` method takes a shelterId and returns the available information on the shelter from the Petfinder database.

One shelter that's very dear to me is [Seattle Area Feline Rescue](http://www.seattleareafelinerescue.org/), the shelterId of which is 'WA40'. Let's return the information available on the rescue.

```{r}
safr <- pf$shelter.get('WA40')
safr
```

### Getting pet records from a shelter

Let's say we are interested in finding the available pets at a particular shelter, we could use the `shelter.getPets()` method to quickly find this information. We can get a `data.frame` of the cats (SAFeR is a cat-only rescue) currently listed on Petfinder that are at the shelter.

```{r}
safr.pets <- pf$shelter.getPets('WA40')
```

### Finding shelters that have a particular animal breed

The `shelter.listByBreed()` method allows one to find shelters that have a particular breed of animal, which may be useful for users interested in finding rarer breeds of animal. The `breed` argument must match a breed listed in the Petfinder database, which can be easily extracted by using the `breed.list()` method. We already pulled the available cat breeds earlier in the vignette, so we can use that to select a breed of cat to search.

The Abyssinian is a beautiful and wonderfully friendly breed of cat, so let's search for some shelters that have Abyssinians listed on Petfinder.

```{r}
aby_shelters <- pf$shelter.listByBreed('cat', 'Abyssinian')
aby_shelters
```
