---
title: "PetfindeR Introduction Part Two"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with PetfindeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction to `PetfindeR` Part Two

The first vignette introduced and explored the basic usage of the `PetfindeR` library. In this vignette, we take a quick look at some of the additional uses of the library and its methods to extract data from the Petfinder database. Before getting started, it is worthwhile to recall the limits imposed by Petfinder when interacting with its API. The following restrictions are copied from [Petfinder's API documentation](https://www.petfinder.com/developers/api-docs#restrictions):

* Total requests per day: 10,000
* Records per request: 1,000
* Maximum records per search: 2,000

Therefore, a single call cannot exceed 2,000 records, and there is a maximum of 1,000 records per request. Knowing this information, we can construct our calls to the API to extract the maximum amount of information without exceeding the limitations imposed by Petfinder.

## Getting Started

After `PetfindeR` is installed, it can be loaded using the `library()` function. We initialize the `Petfinder` R6 class object using a key received from Petfinder. You can easily obtain a key by creating an account on Petfinder on the [site's developers page](https://www.petfinder.com/developers/api-key). Once the key is obtained, we use that as an argument when initializing the class. We can now begin to extract data from the Petfinder API!

```{r}
library(PetfindeR)
key <- Sys.getenv('PETFINDER_KEY')
pf <- Petfinder(key)
```

## Paging Results

The Petfinder API pages results to reduce the stress on the API when extracting a large number of records. Paging is the only way to exceed the API's limitation of 1,000 records per request. For example, to get 2,000 animal records located in Washington State (and beyond), we could set the `count` parameter in the `pet.find` method to $500$ and the `pages` parameter to $4$. We could also change the `count` and `pages` parameters to $1,000$ and $2$, respectively, which would achieve the same result.

```{r, eval=FALSE}
paged <- pf$pet.find(location = 'WA', count = 500, pages = 4, return_df = TRUE)
paged2 <- pf$pet.find(location = 'WA', count = 1000, pages = 2, return_df = TRUE) # Returns the same as above
```

The `PetfindeR` library will check to make sure the `count` parameter doesn't exceed 1,000 to avoid wasting an API call.

```{r, error=TRUE}
count.too.high <- pf$pet.find(location = 'WA', count = 1001, return_df = TRUE)
```

The library will also check to make sure that the input `pages` parameter and `count` will not exceed 2,000.

```{r, error=TRUE}
max.record.exceeded <- pf$pet.find(location = 'WA', count = 500, pages = 5, return_df = TRUE)
```

## Offsetting Results

Four of the available methods (`pet.find`, `shelter.find`, `shelter.getPets`, and `shelter.listByBreed`) in the Petfinder API offer an `offset` parameter which can be used to skip the first $n$ results from the API. Behind the scenes, the `offset` parameter is used to page results if the `pages` parameter is specified, but it can also be used to avoid returning the same data when iterating calls. For example, the below two calls using the `pet.find` method would, in essence, return the first 100 results should the two results be appended.

```{r, eval=FALSE}
no.offset <- pf$pet.find(location = 'WA', count = 50, return_df = TRUE) # Get the first 50 results
offsetted <- pf$pet.find(location = 'WA', offset = 50, count = 50, return_df = TRUE) # Skip the first 25 results and get the next 50
```

The above two calls could also be written as the below.

```{r, eval=FALSE}
paged <- pf$pet.find(location = 'WA', pages = 2, count = 50, return_df = TRUE)
raw_count <- pf$pet.find(location = 'WA', count = 100, return_df = TRUE)
```

## Conclusion

This short vignette explored some of the additional functionality of several of the methods available in the Petfinder API and `PetfindeR` library for further customizing the number of records and their location in the database using the `offset` parameter. Future vignettes will also explore some possible use cases of extracting and analyzing the data available in the Petfinder database.